---
title: "ggplot2 Fundamentals (Part 2): Structure, Graph Types, and Better Visualisations"
date: 2026-03-21
description: "A beginner-friendly guide to ggplot2. Learn the basic structure, how to choose the right graph based on your data, and how to make your visualisations more readable."
categories: 
 - ggplot2
 - Data Visualisation
format:
  html:
    include-after-body: blog-footer.html
toc: true
---

This is a three-part series. Other series can be viewed here:

-   [ggplot2 Fundamentals (Part 1): Choosing the Right Graph for Your Data](https://reinasdatastudio.github.io/webpage/blog/posts/18-03-26-choosing-the-right-graph.html)

-   [ggplot2 Fundamentals (Part 3): Useful ggplot Techniques](https://reinasdatastudio.github.io/webpage/blog/posts/25-03-26-ggplot_part3.html)

------------------------------------------------------------------------

```{r include=FALSE}
library(ggplot2)
```

When creating graphs in **R**, one of the most widely used tools is **ggplot2**.

However, many beginners feel overwhelmed when they first encounter ggplot. The code can look long, and it’s often unclear what each part is doing.

The good news is that the core idea behind ggplot is actually very simple.

In this article, we will cover:

-   The basic structure of ggplot
-   How to choose the right type of graph
-   How to make your graphs easier to read

This article combines key ideas from Parts 2–4 of my ggplot series in Japanese (R seems to be a lot less common in Japan, hence my ggplot blog posts are more detailed in Japanese).

# Basic Structure of ggplot

Most ggplot code follows this structure:

```         
data + aesthetics + geometry
```

In other words:

```         
data + how it looks + type of graph
```

In code, this looks like:

```{r eval=FALSE}
ggplot(data, aes(x, y)) +
  geom_point()
```

Here, you are specifying two main things:

-   Which data to use
-   What type of graph to create

Even with just these three components, you can create a basic plot.

## Example: Scatterplot

Using the built-in `iris` dataset:

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point()
```

This creates a scatterplot where:

-   `Sepal.Length` is mapped to the x-axis
-   `Petal.Length` is mapped to the y-axis

# Choosing the Right Type of Graph

One of the most important steps in data visualisation is choosing the correct graph.

This depends on the <strong>**type of variables**</strong> you are working with.

There are two main types:

-   <strong>**Continuous variables**</strong> (measured values, e.g. length, weight)
-   <strong>**Discrete variables**</strong> (categories or counts)

For example, in the `iris` dataset:

-   `Sepal.Length` and `Petal.Length` are continuous
-   `Species` is discrete

## Common Graph Choices

| Data Type                | Graph        | ggplot             |
|--------------------------|--------------|--------------------|
| Discrete (1 variable)    | Bar chart    | `geom_bar()`       |
| Continuous (1 variable)  | Histogram    | `geom_histogram()` |
| Continuous (1 variable)  | Density Plot | `geom_density()`   |
| Continuous (2 variables) | Scatterplot  | `geom_point()`     |
| Discrete + Continuous    | Boxplot      | `geom_boxplot()`   |

Choosing the right graph depends on what you want to explore:

-   Distribution → histogram
-   Relationship → scatterplot
-   Comparison across groups → boxplot

# Making Your Graphs More Readable

Even with the same data, small changes can significantly improve readability.

Let’s start with a simple scatterplot:

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) + 
  geom_point()
```

This shows the data, but it can be improved.

## Adding Color

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point()
```

This helps distinguish different groups.

## Improving Labels

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point() +
  labs(title = "Sepal Length vs Petal Length",
       x = "Sepal Length (cm)",
       y = "Petal Length (cm)")
```

Adding labels makes the graph easier to interpret.

## Using Themes

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point() +
  theme_minimal()
```

Themes help simplify the visual appearance.

# Key Takeaways

-   ggplot is built on a simple structure: **data + aesthetics + geometry**
-   Choosing the right graph depends on: **your variable types (continuous vs discrete)**
-   Small improvements like color, labels, and themes can make your graphs much clearer

# Final Thoughts

The most important step is not writing code - it’s understanding your data.\
\
Once you have that, ggplot becomes much easier to use.

If you’re just getting started, focus on:

-   Understanding your data
-   Choosing the right graph
-   Improving clarity step by step

These skills will make a big difference in how effectively you communicate your analysis.

## Reference Materials

This page explains all about ggplot2 code in depth:

<https://rstudio.github.io/cheatsheets/html/data-visualization.html>
